---
title: "Beginning Visualization and Data Exploration"
author: "kira"
output:  
  html_document:
     toc: true
     toc_float: true
     theme: cerulean
     highlight: tango
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T
                      ,message = F
                      ,warning = F
                      ,dpi=300
                      ,fig.width=5.5
                      ,fig.height=4.8)
library(tidyverse)
options(tigris_use_cache = TRUE)

thme <- 
  list(
    theme_light()
    ,theme(text = element_text(size = 9))
  )

```

(Using [COVID-19 data](https://data.cityofnewyork.us/Health/COVID-19-Daily-Counts-of-Cases-Hospitalizations-an/rc75-m7u3) from NYC open data.)

This short document is meant to demonstrate some process for loading and exploring a dataset and shaping some research questions based on the initial exploration.

This document is a "markdown," which is a document that can be created in R that can combine code, results, and plain-language writing all in one.

# Initial load & explorations

The following block of code will load and do some initial processing of COVID-19 data from NYC OpenData.

```{r, results='hide'}

lnk <- 'https://raw.githubusercontent.com/nychealth/coronavirus-data/master/trends/data-by-day.csv'

cd <- vroom::vroom(lnk)

cd <- cd %>% rename_with(tolower)

# lubridate
library(lubridate)

cd <- cd %>% 
  rename(date = date_of_interest) %>% 
  mutate(date =
           mdy(date)
           )

#cd

cd %>% arrange(desc(date)) %>% glimpse() 

# regular expressions -- can be useful, but won't be a focus in this class. See https://r4ds.hadley.nz/regexps
boro.prefix.regx <- c('^bx|^bk|^mn|^qn|^si')

# create a by-boro dataset (don't worry if you don't understand this whole block of code)
cdbb <- cd %>% 
  select(date,
         matches(boro.prefix.regx)
         ) %>% 
  pivot_longer(
    cols = matches(boro.prefix.regx)
    ,names_sep = 2
    ,names_to = c('boro','var')
  ) %>% 
    mutate(var = 
             trimws(
               gsub('_', ' ', var)
    ))

cdbb

cdbb$date %>% min()

```

# Starting out!

A table with daily averages is shown below:

```{r}

# check what variables are available
cdbb %>% count(var)

# filter to counts
ccounts <- cdbb %>% 
  filter(grepl('count$', var)) %>% 
  mutate(var =
           gsub(' count$', '', var))

# calculate means by outcome (& format!)
means.by.outcome <- ccounts %>% 
  group_by(var) %>% 
  summarise(value = 
              scales::comma(mean(value)))

# output! `knitr::kable` makes output prettier in a markdown.
knitr::kable(means.by.outcome)

```

## Initial questions

Where can we go from here? what questions may be raised by this category?

* disaggregate by borough?
* patterns over time?

What questions might we have?

* What constitutes a "Covid-19 death"? Does covid have to be a leading factor or can it be a contributing one? Is there much ambiguity?
* What is a "probable case"? How do they differ from "cases"? How can we approximate likely actual cases -- should we add them together? 
 
 
This dataset is relatively well documented! So we can look at that documentation

  * But note: the documentation is outdated. We'll have to poke around dataset to see how categories actually fit together. Unfortunately, this can be pretty common for all but the best-maintained datasets.

# Disaggregating by Borough

```{r}

full.boro.nms <- 
  tibble(
    abv = Hmisc::Cs(si, qn, mn, bx, bk)
    ,boro.nm = Hmisc::Cs('Staten Island', 'Queens', 'Manhattan', 'Bronx', 'Brooklyn')
  )

ccounts %>% 
  filter( !grepl('case',var) ) %>%
  group_by(boro, var) %>% 
  summarise(value = sum(value)
            ) %>% 
  left_join(full.boro.nms
            ,by = c('boro' = 'abv')) %>% 
  ggplot(
    aes( y = boro.nm
        ,x = value
        ,fill = var)
  ) +
  geom_col() +
  scale_fill_manual(
     NULL
    ,values = visaux::jewel.pal()
  ) +
  scale_x_continuous(
    name = NULL
    ,labels = scales::comma
  ) +
  scale_y_discrete(
    name = NULL
    ,limits=rev
  ) +
  theme_light() +
  labs(
     title = 'Covid Deaths and Hospitalizations in NYC, by Boro'
  )

```

*How could we go further again?*

* Adding caption for source!
* Adding takeaway title!
* Normalizing by population. Ratios of hospitalization/deaths : population?
* Sorting boroughs more logically.. or mapping!

What other questions may be raised? How might the ambiguities in the data affect how we should interpret those findings?

# Data Over time

```{r, results='hide'}

# could  use the below to re-calc moving averages
#library(zoo)

# again, remind what variables were included
cdbb %>% count(var)

# use 7-day moving averages already included in the data
mavgs <- cdbb %>%
  filter(grepl('7day avg$', var)) %>% 
  mutate(var = 
           gsub(' count 7day avg', '', var)) %>% 
  rename(ma = value) %>% 
   left_join(full.boro.nms
            ,by = c('boro' = 'abv')
            ) 

tmp <- 
  mavgs %>%
  filter(grepl('all case|hospitaliz|death', 
               var)) %>% 
  pivot_wider(
     names_from = var
    ,values_from = ma
  ) %>% 
    rename_with(make.names) %>% 
    mutate(death.or.hospitalized =
             hospitalized + death
           ,deathorhosp2case.ratio =
             death.or.hospitalized /
             all.case
           )  

# quick peek
tmp$deathorhosp2case.ratio %>% summary()

tmp %>% 
  filter(date > ymd('2023-01-01')
         ) %>% 
  ggplot(
    aes( x = date
        ,y = deathorhosp2case.ratio
        ,color = boro.nm
        )) +
  geom_path(
    linewidth = .4) +
  scale_color_manual(
    NULL,
    values = visaux::jewel.pal()
  ) +
  scale_x_date(
     NULL
    ,limits = c(
      ymd('2023-01-01'),
      max(tmp$date)
    )
  ) +
  scale_y_continuous(
    'Ratio of Cases : Deaths or Hospitalizations'
  ) +
  theme_light(
  ) +
  labs(
    title = 'Ratio of Cases to Deaths or Hospitalizations, 2023'
    ,subtitle = '(Higher ratios mean fewer deaths or hospitalizations per COVID incident)'
    ,caption = 'Source: COVID-19 Daily Counts, NYC Opendata, moving 7-day averages.')
  

```

* How could we interpret this graph? How can we frame or posit findings from it?

* What additional questions might it raise? How could we begin to explore those?


# Add'l example NYC openData sets

* [Causes of deaths](https://data.cityofnewyork.us/Health/New-York-City-Leading-Causes-of-Death/jb7j-dtam)
* [Vehicle Crashes](https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95)
* [Covid Counts](https://data.cityofnewyork.us/Health/COVID-19-Daily-Counts-of-Cases-Hospitalizations-an/rc75-m7u3)
* [Police reports]( tohttps://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Historic/qgea-i56i)

